Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-02 pkg/paper/Paper/draft.Rnw
updated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@110 b679b2f7-6539-45d0-a605-5ddcb976cde9

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution, which we implemented in a new \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} 
  \citep{R} package called \pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]}. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change towards negative skewness.
}

@@

Ever since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area (as developed in: \citet{Mundell:1961})\footnote{
For an overview, the interested reader is referred to \citet{Dellas:2009}}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate: this regards the question about short-run and steady-state
inflation uncertainty (in context to inflation expectations) dealt with in
\citet{cap}, or the structural convergence of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

\readme{perhaps skip this paragraph?} The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.
Any new state that wants to join the EMU is first obliged to fulfill the
Maastricht criteria, which, besides setting rules for government debts and
interest rates, also requires the participation in the ERM~II (Exchange Rate
Mechanism~II) for two years, while the exchange rates towards the euro are not
allowed to cross the nominal band.

\clearpage
\cite{emerson} put emphasize on the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism.  Failing to stabilize inflation leads to
severe economic problems. Throughout the literature, there is still a
considerable degree of uncertainty as to what extent the introduction of the
euro, or monetary unions in general, affect the inflation rate. The EMU requires
a country to possibly alter a number of economic policies and institutions,
which may require significant short term adjustments.

Given that a country experienced quite volatile inflation rates, its efforts to
meet the convergence criteria are likely to lead to an alteration at least in
the mean (since this was one of the requirements) and possibly in the volatility
of their respective inflation rates as well.

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this  by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries. The framework developed in this paper makes it possible to test for any 
significant changes of a countries inflation dynamics and thus helps interpreting 
the effects of the EMU. We focus on changes in the first three moments of these 
inflation rates, which we assume to be distributed according to a generalized logistic 
distribution (GL). The GL distribution is a suitable one due to its flexibility to 
capture nonsymmetry and because of its fatter tails. For this purpose, we develop a 
new estimation approach building on previous research by \citet{Zeileis+Hornik:2007}.

This \fixme{Jesus: sell method better, why this one and not another} framework
allows us to assess the evolution of mean, variance and skewness over time. The
skewness parameter turns out to be quite important to measure alterations in the
long run inflation dynamics, as it changes considerably over the time of the
analysis. This trend would be all but invisible using a normal distribution.  
Our estimation technique enables the simultaneous estimation of breaks in the
mean (which is important in the context of the Maastricht criteria), the
variance (which should be low for the price mechanism to be efficient) and the
skewness (where a change in skewness is evidence of a shift of the distribution
towards a new ``center of gravity'') of the inflation rates of interest. This
than can be used to search for answers regarding the question about the effect
of the EMU on the inflation dynamics of its member states or more general to
investigate the effects of external shocks.

The \fixme{Jesus: maybe comment on results here} remainder of this paper is
structured as follows: Section~\ref{sec:data} presents the data,
Section~\ref{sec:model} presents the model and the estimation techniques used,
Section~\ref{sec:slovenia} illustrates our approach using Slovenia as an
example, Section~\ref{sec:results} presents the results and
Section~\ref{sec:conclusion} concludes.

@@

<<data>>=
data("hicps", package = "glogis")
eu <- na.omit(hicps[, "EU"])
hicps <- hicps[, colnames(hicps) != "EU"]
@

@@

The countries in this sample can be divided into three different groups: (1)~euro
countries -- Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia,
Spain; (2)~EU members not participating in the ERM~II -- Czech Republic,
Hungary, Poland, United Kingdom, Sweden; (3)~Denmark which stands on its own as a
member of the EU and the ERM~II, but not yet of the EMU. 
Latvia, Lithuania, Bulgaria, Romania are excluded due to data scarcity.
Cyprus and Malta are not included since they are very small economies.

Seasonal adjustment is necessary due to the harmonization of the treatment
of sales prices\footnote{Inclusion of sales prices in the HICP was demanded by
a Commission regulation \citep[see][for details]{eurlex}.} in the HICP, resulting in
changes of the seasonal inflation patters over time within several countries
and also between the countries. See \cite{pltm} for a detailed discussion.

@@

Hence, a likelihood-based model is adopted that can also incorporate stylized
facts for return data, such as heavy tails and skewness. Here, the seasonally
adjusted returns $y_i$ ($i = 1, \dots, n$) are assumed to come from a
generalized logistic (GL) distribution \citep[see][]{johnson} with
three-dimensional parameter $\phi_i = (\theta_i, \sigma_i, \delta_i)^\top$ for
location~$\theta_i$, scale~$\sigma_i$, and shape~$\delta_i$ at time $i$. Structural change
techniques are then employed to check whether the parameter vector $\phi_i$
remains constant over time and, if that is not the case, when and how the
parameters change. This framework allows to trace breaks in the
evolution of mean, variance and skewness of the inflation series which may
possibly be linked to underlying regime changes, e.g., in monetary policy.

For estimation of the parameters, a (quasi-)maximum likelihood framework is
adopted and the observations are assumed to be (approximately) independent,
i.e., potential autocorrelation is treated as a nuisance parameter and not
explicitly incorporated in the model. The motivation for this is to focus on the
shifts in mean, variance, and skewness -- and it turns out to be a useful model
for the data under investigation because only negligible amounts of residual
autocorrelation remain after incorporation of the structural breaks. If the
focus of the analysis were the dynamics in the autocorrelation, higher frequency
data would be required, e.g., in order to apply techniques such as those
expounded by \cite{andreou}.\footnote{For example, it would be conceivable to
obtain high-frequency data for certain years from the Billion Prices Project of
the \cite{bpp}; however, this is not pursued here.}

@@

The logistic distribution is often used in econometrics in the
context of income and growth models due to its fatter tails compared to the 
normal distribution -- it has also been applied to the modeling of
inflation rates by \citet{batchelor}. Its generalization -- which additionally
allows for asymmetries -- has been applied to analyze risk \citep[e.g., by][]{tolikas}
but has not yet been applied -- to the best of our knowledge -- to inflation rates.
However, it will prove to be a rather simple model that fits HICP inflation rates
quite well.

The GL distribution in the parameterization employed here is also used in \citet{Shao:2002}
% The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{Abberger:2002}
%\citet{johnson},
and has the log-density function\footnote{Which is similar to the type~1 GL distribution as defined in \citet{johnson}.}
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$ and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

The modeling strategy proposed in Section~\ref{sec:model} is applied to each of the
countries included in the data set introduced in Section~\ref{sec:data}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the segment-specific
moments (mean, variance, skewness) implied by the corresponding parameter
estimates. Furthermore, entry dates for ERM/ERM~II and the euro are included as
additional information. For ease of interpretation, countries with similar patterns of
change are grouped together in Table~\ref{tab:results1} (and ordered alphabetically
within each group).

If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at the 5\%
level. In significant cases, the number of breakpoints was selected via the LWZ criterion
and subsequently all parameters (breakpoints and segment-specific GL parameters) were
estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation
in the empirical scores after inclusion of the breakpoints (if any).

@@

\noalign{\medskip}
\multicolumn{8}{l}{\emph{European Union}} \\
\hline
<<moments-table6, results=tex>>=
writeLines(unlist(lapply(22, format_moments)))
@

@@

\subsection{The countries in detail}

For Table~\ref{tab:results1}, we grouped the countries according to either their
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries had and still have 
a very low inflation rate in terms both of mean and variance and their inflation dynamics
did not change. This does make economic sense since the ECB is build after the tradition of
the Deutsche Bundesbank and the ECB did not -- until recently -- much diverge from the 
monetary policy of the former.

Austria, Finland and France adjusted to the requirements of the EMU during the
first phase of convergence. These are also the countries with the lowest
inflation rates in both mean and variance after the adjustment. In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS (Portugal, Ireland, Italy, 
Greece and Spain) countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successfull in reducing its inflation rate, whereas Luxembourg
experienced a doubling of its former -- very low -- inflation rate with the
variance rising by a factor of 14 with a strong shift in skewness towards the
left. It is possible that in the case of Luxembourg, this still reflects the
adjustment due to the treatment of sales periods, even after seasonal adjustment
of the original series.

The forth subgroup consists of two of the PIIGS  countries, namely Ireland and Spain.  
Both show a break in the summer of  2008  following the financial crisis and the severe 
problems in both countries concerning   the real estate situation. In both countries this  
had a strong deflationary impact,  with  Ireland actually showing negative inflation rates 
for an extended period of time;   in Spain, which was affected by a housing bubble of  
roughly the same size as Ireland,  the housing prices did not adjust so abruptly and  
in such a degree. This deflationary  trend, which for a time affected all countries,  
can be seen in Figure~\ref{fig:glmean},  which depicts a rolling mean estimation,  
where the mean is calculated using the first  moment of a fitted GL distribution.  
There is evidence of a decline in mean inflation,  which returns to its previous 
path roughly a year later.

\clearpage
The next group consists of two countries that decided not to be part of the
European Monetary System\footnote{If we ignore the short participating time of
both countries in the EMS.}, the United Kingdom and Sweden. Both show a clear
break in the early 90s. These can be traced back to economic crisis, the currency
crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and
the banking crisis in Sweden during the early 90s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
later part of the 90s (Slovenia is an exception with an increase in variance
after the break). Most of them experienced a break in 1997/1998, with Poland,
Slovakia and Slovenia somewhat later. The break during the late 90s is a result
of these countries efforts to curb in inflation by decreasing the growth of the money 
supply and indicates the time when these countries actually overcame the biggest
transitionally shocks on their way towards free market economies. Slovakia is
very similar to Slovenia, as its inflation dynamics too changed roughly one year
prior to its entry into ERM~II.

\subsection{Overview over the general trend}

The disparity of the national inflation rates within the EMU is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation
differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity
across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU
countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that
convergence over time of inflation dynamics was only partly observable.  In a paper studying core inflation and using an aggregated euro area inflation
rate, \citet{morana} finds three regimes (roughly 1980--1984, 1984--1993 and 1993--2000) governing the core inflation rate.

Concerning inflation differentials, Figure~\ref{fig:glmean} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months),
where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month,
the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month-on-month increase at any one time. The
red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the
short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. The general trend also suggests, that the inflation rates
converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see
an increase in mean from
$\Sexpr{round3(eu_mom[1, 1])}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
$\Sexpr{round3(eu_mom[2, 1])}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
$\Sexpr{round3(eu_mom[3, 1])}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) following the financial crises.
Whereas variance has increased considerably over time from
$\Sexpr{round3(eu_mom[1, 2])}$ over $\Sexpr{round3(eu_mom[2, 2])}$ to $\Sexpr{round3(eu_mom[3, 2])}$.
Regarding skewness, we witness a movement towards negative skewness from
$\Sexpr{round3(eu_mom[1, 3])}$ over $\Sexpr{round3(eu_mom[2, 3])}$ to $\Sexpr{round3(eu_mom[3, 3])}$.

@@

\fixme{need to note somewhere} Rapid changes that persist only for very short amounts of time
are not captured well. Specifically, short-lasting changes in the mean, may simply be attributed
to increased variance. For example, the mean inflation in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the year after that (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from rolling GL fits (red)
over time for all countries.}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

Regarding the general trend, Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany,
Luxembourg and Netherlands), the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments.

@@

In the core-zone, we see a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[4]}, with no change in the median inflation rate, which is due to the
fact that there is no break after \Sexpr{yrs[2]} for these countries; however,
the interquartile range decreased by almost 20\%. The variance increased
slightly (but with a 15\% decrease in interquartile range), whereas skewness
stayed symmetric in the median, with almost no change in the interquartile
range.

The PIIGS countries decreased inflation rates as well with a strong change of skewness 
towards more negative values. The interquartile range of their inflation variance increased
considerably, which is mostly due to Spain and Ireland and their strong reaction
to the financial crisis.

The other countries were successfull in decreasing all three moments, but they
are still higher than the euro-zone countries. Most importantly, the
interquartile range for all moments decreased considerably showing a clear
convergence towards the core-zone countries.

>>>>

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change
  towards negative skewness.

}

@@

\readme{perhaps skip this paragraph?} The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence, and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.

Since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area \citep[as developed in:][]{Mundell:1961}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate. This concerns the issue of short--run and steady--state
inflation uncertainty -- in context to inflation expectations -- dealt with in
\citet{cap}, or the degree of similarity of short--run dynamic properties of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

%Any new state that wants to join the EMU is first obliged to fulfill the
%Maastricht criteria, which, besides setting rules for government debts and
%interest rates, also requires the participation in the ERM~II (Exchange Rate
%Mechanism~II) for two years, while the exchange rates towards the euro are not
%allowed to cross the nominal band.

Throughout the literature, there is still a
considerable degree of uncertainty as to what extent the introduction of the
euro, or monetary unions in general, affect the inflation rate.

However, a deeper inquiry into the effects of monetary unions on the inflation rates of its member
countries is in need of a method able to
capture changes in inflation dynamics. We contribute to this field of research by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates. The framework utilized in this paper allows testing for any 
possible alternations  and thus helps interpreting 
the effects of the EMU. We focus on changes in the first three moments of these 
inflation rates, which we assume to be distributed according to a generalized logistic 
distribution (GL). For this purpose, we build on previous research by \citet{Zeileis+Hornik:2007}.

This \fixme{describe method} framework allows us to assess the evolution of mean, variance, and skewness over time.
Our estimation technique enables the simultaneous estimation of breaks in these moments. 
The skewness parameter turns out to be quite important to measure alterations in the
long run inflation dynamics, as it changes considerably over the time of the
analysis. This trend would be all but invisible using a normal distribution.

\fixme{comment on results here} Overall, we find four periods during which structural breaks
happened. Two of them in the 90s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
tendency towards a lower mean inflation rate, combined with a decreasing difference between
individual inflation rates.

The  remainder of this paper is
structured as follows: Section~\ref{sec:data} presents the data,
Section~\ref{sec:model} presents the model and the estimation techniques used,
Section~\ref{sec:slovenia} illustrates our approach using Slovenia as an
example, Section~\ref{sec:results} presents the results, and
Section~\ref{sec:conclusion} concludes.

@@

<<data>>=
data("hicps", package = "glogis")
eu <- na.omit(hicps[, "EU"])
## do not include EU for figures
hicps <- hicps[, colnames(hicps) != "EU"]
@

@@

The countries in this sample can be divided into three different groups: (1)~euro
countries -- Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia,
Spain; (2)~EU members not participating in the ERM~II -- Czech Republic,
Hungary, Poland, United Kingdom, Sweden; (3)~Denmark which stands on its own as a
member of the EU and the ERM~II, but not yet of the EMU. 
Latvia, Lithuania, Bulgaria, Romania are excluded due to data scarcity.
Cyprus, and Malta are not included since they are very small economies.

Seasonal adjustment is necessary due to the harmonization of the treatment
of sales prices\footnote{Inclusion of sales prices in the HICP was demanded by
a Commission regulation \citep[see][for details]{eurlex}.} in the HICP, resulting in
changes of the seasonal inflation patters over time within several countries
and also between the countries. See \citet{pltm} for a detailed discussion.

@@

Hence, a likelihood-based model is adopted that can also incorporate stylized
facts for return data, such as heavy tails and skewness. Here, the seasonally
adjusted returns $y_i$ ($i = 1, \dots, n$) are assumed to come from a
generalized logistic (GL) distribution \citep[see][]{johnson} with
three-dimensional parameter $\phi_i = (\theta_i, \sigma_i, \delta_i)^\top$ for
location~$\theta_i$, scale~$\sigma_i$, and shape~$\delta_i$ at time $i$. Structural change
techniques are then employed to check whether the parameter vector $\phi_i$
remains constant over time and, if that is not the case, when and how the
parameters change. This framework allows to trace breaks in the
evolution of mean, variance, and skewness of the inflation series which may
possibly be linked to underlying regime changes, e.g., in monetary policy.

For estimation of the parameters, a (quasi-)maximum likelihood framework is
adopted and the observations are assumed to be (approximately) independent,
i.e., potential autocorrelation is treated as a nuisance parameter and not
explicitly incorporated in the model. The motivation for this is to focus on the
shifts in mean, variance, and skewness -- and it turns out to be a useful model
for the data under investigation because only negligible amounts of residual
autocorrelation remain after incorporation of the structural breaks. If the
focus of the analysis were the dynamics in the autocorrelation, higher frequency
data would be required, e.g., in order to apply techniques such as those
expounded by \citet{andreou}.\footnote{For example, it would be conceivable to
obtain high-frequency data for certain years from the Billion Prices Project of
the \citet{bpp}; however, this is not pursued here.}

@@

The logistic distribution is often used in econometrics in the
context of income and growth models due to its fatter tails compared to the 
normal distribution -- it has also been applied to the modeling of expected
inflation rates by \citet{batchelor}. Its generalization -- which additionally
allows for asymmetries -- has been applied to analyze extreme risks in the context of stock markets \citep[e.g., by][]{tolikas}
but has not yet been applied -- to the best of our knowledge -- to inflation rates.
However, it will prove to be a rather simple model that fits HICP inflation rates
quite well.

The GL distribution in the parameterization employed here is also used in \citet{Shao:2002}
% The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{Abberger:2002}
%\citet{johnson},
and has the log-density function\footnote{Which is similar to the type~1 GL distribution as defined in \citet{johnson}.}
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$, and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

\readme{questions and challenges} The aim of this paper is to test for structural breaks in the dynamics of the inflation rates. Making use of the method developed in Section~\ref{sec:model},  the effect of the EMU upon national inflation rates can be investigated. Apart from the countries of the euro--zone, the inflation dynamics of the other European countries are also of interest. 
This issue is closely related with the question about the convergence of the national inflation rates of the EMU members towards some common mean. Many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials in parts of the sample.\readme{perhaps skip this one??} \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity in terms of average inflation and its degree of persistence.

\readme{methods, approach} To explore these issues, we make use of three different approaches. First, in Section~\ref{sec:bpresults}, the estimation technique proposed in Section~\ref{sec:model} is applied to each of the countries included in the data set introduced in Section~\ref{sec:data}. Table~\ref{tab:results1} reports the estimated breakpoints along with the segment-specific moments (mean, variance, skewness) implied by the corresponding parameter
estimates. Furthermore, entry dates for ERM/ERM~II and the euro are included as additional information. For ease of interpretation, countries with similar patterns of
change are grouped together and ordered alphabetically within each group.
If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at the 5\% level. In significant cases, the number of breakpoints was selected via the LWZ criterion and subsequently all parameters (breakpoints and segment-specific GL parameters) were estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation in the empirical scores after inclusion of the breakpoints (if any).

Section~\ref{sec:aggregate} deals with the convergence issue. Table~\ref{tab:results2} gives an overview over the data aggregated into three different country groups: the core--euro countries (Austria, Belgium, Finland, France, Germany, Luxembourg, and Netherlands), the PIIGS countries (Portugal, Ireland, Italy, Greece, and Spain), and the other countries. This is done by showing mean, variance, and skewness over time and taking possible break points in the data into account.

Finally, Figure~\ref{fig:glmean} in Section~\ref{sec:fcrisis}, depicts the inflation rates and a rolling mean estimation. This is another way of investigating into a possible convergence pattern of national inflation rates and is especially useful to depict the influence of the financial crisis.

@@

\noalign{\medskip}
\multicolumn{8}{l}{\emph{Euro--area}} \\
\hline
<<moments-table6, results=tex>>=
writeLines(unlist(lapply(22, format_moments)))
@

@@

\subsection{Breakpoint estimation results}\label{sec:bpresults}

\readme{changed to: 1) individual countries (table 1), 2) EU (table 1), 3) groups of countries (table 2), 4) crisis (fig.3)}
For Table~\ref{tab:results1}, we grouped the countries according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries had and still have 
a very low inflation rate in terms of both mean and variance and their inflation dynamics
did not change. This does make economic sense since the ECB is modeled after 
the Deutsche Bundesbank and the ECB did not -- until recently -- much diverge from the 
monetary policy of the former. Thus the EMU did have no effect since the policies in these countries were
already much in line with Germany and thus the ECB.

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS  countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successfull in reducing its inflation rate, whereas Luxembourg
experienced a doubling of its former -- very low -- inflation rate with the
variance rising by a factor of almost 14 with a strong shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successfull in reducing mean inflation rates. In this group, as well
as in the previous, the EMU adjustment led to a lower mean inflation rate.

The forth subgroup consists of two of the PIIGS  countries, namely Ireland and Spain.  
There is a break in both countries in the summer of  2008  following the financial crisis and the severe 
problems concerning the real estate situation. In Ireland as well as in Spain, this  
had a strong deflationary impact, with  deflation in Ireland over an extended period of time. 
In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland,  the housing prices did not adjust so abruptly and  
in such a degree.

%This deflationary  trend, which for a time affected all countries,  
%can be seen in Figure~\ref{fig:glmean},  which depicts a rolling mean estimation,  
%where the mean is calculated using the first  moment of a fitted GL distribution.  
%There is evidence of a decline in mean inflation,  which returns to its previous 
%path roughly a year later.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and in that way also not to participate in the EMU, i.e. the United Kingdom and Sweden. Both show a clear
break in the early 90s. These can be traced back to economic crisis, the currency
crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and
the banking crisis in Sweden during the early 90s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
later part of the 90s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998. This is a result
of these countries efforts to curb in inflation by decreasing the growth of the money 
supply and indicates the time when they actually overcame the biggest
transitionally shocks on their way towards free market economies. In Poland, this took two more
years. Slovakia and Slovenia are special cases: for once, they started with lower mean
inflation values and on the other hand, their inflation dynamics  both changed roughly one year
prior to their entry into the ERM~II.

@@

In the case of the inflation rate of the euro--area, we witness three different regimes. One until 
1999 with both a low mean and variance, the next shortly after the beginning of the third and final stage of
monetary integration, accompanied with a significant increase in the mean. The overheating of the economy that
culminated in the financial crisis shows itself in a big increase in variance and a lower mean inflation, reflecting
the negative inflation rates is some euro--area countries.

%There is an increase in mean from
%$\Sexpr{round3(eu_mom[1, 1])}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
%$\Sexpr{round3(eu_mom[2, 1])}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
%$\Sexpr{round3(eu_mom[3, 1])}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) shortly before the financial crises.
%

\clearpage
\subsection{Convergence on an aggregate level}\label{sec:aggregate}

@@

Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone, the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments.

The time periods are chosen according to four fundamental change periods. Namely, the early 90s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 90s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. And also the period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the two countries with the most visible effect of the financial crisis (Ireland and Spain).

In the core-zone, there is a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]}. However, the interquartile range decreased by almost 20\%. The variance increased
slightly, but also with a 15\% decrease in interquartile range. The skewness
stayed symmetric in the median, with almost an 8\% increase in the interquartile range.

The PIIGS countries significantly decreased the median inflation rates from \Sexpr{yrs[1]} to \Sexpr{yrs[4]}. 
Whereas the interquartile range was very tight in \Sexpr{yrs[1]}, it increased in \Sexpr{yrs[2]} and again in \Sexpr{yrs[4]}.
However, the median values are much smaller. The variance increased from \Sexpr{yrs[1]} to \Sexpr{yrs[2]} but did not change much 
in \Sexpr{yrs[4]}. The interquartile range did not change much too. Median skewness changed considerable 
from highly positive to negative values. This is mostly due to the changes associated with the financial crisis, that was
responsible for deflationary shocks. The interquartile range of skewness decreased considerably.
